Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-01-25 draft.tex
More small changes, adjust formatting and find a proper reference to replace [2]

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare decay FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, note that the method we
describe in the following is general, and it applies to arbitrary decay
or scattering processes. We find a previous work \cite{Dighe:1998vk}
that advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays, but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this previous
work by studying the uncertainties that are introduced by mismodelling
of the angular distribution, and by working out a recipe to determine
and unfold detector effects. Moreover, we show
that the method of moments has two major advantages over the usual
approach based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF, see e.g. \cite{Egede:2008uy}.
        The method of moments has no such convergence problems.
    \item The results of likelihood fits suffer from systematic uncertainties
        and systematic bias (see e.g. \cite{Egede:2008uy}), especially in but not limited
        to cases in which the PDFs do not faithfully model the underlying physics. We will
        show that the method of moments does not suffer a very particular type of
        mismodelling that arises from introducing a cutoff in a partial wave expansion.
    \item Using the method of moments, the joint probability distribution of the
        angular observables converges towards a multivariate Gaussian distribution.
        This allows an easy transfer of correlation information from the experiments to
        interested theorists.
\end{enumerate}

@@

Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the distribution of the $\mathrm{pull}$
        of any observables $S$,
        \begin{equation}
            \mathrm{pull_i} \equiv \frac{\est{S} - S}{\est{\sigma}_S}\,.
        \end{equation}
        Here $S$ refers to the true (input) value for the angular observables, $\est{S}$ refers to the
        MC estimate, and $\est{\sigma} \equiv \sqrt{\est{\Sigma}}$ refers to an estimator of the
        variance of the measurement. All pull distributions obtained in our studies can be successfully
        fitted to a gaussian distribution. Due to the large number of studied distributions, we have
        to restrict ourselved to a subset. In our opinion the pull distributions for the observables
        $S_5 \simeq 27\%$ and $S_7 \simeq 2\%$, which are obtained from SM-like $B\to K^*\ell^+\ell^-$
        decays, are good examples. We show their respective pull distributions for \danny{$XXX$} toy studies
        of $200$ events per study in \reffig{pulls}.
    \item We study the absolute uncertainty $\sigma_k(N)$ with respect to the angular observable $S_k$
        as a function of the number of simulated events $N$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \label{eq:unc-on-mean}
            \sigma_k(N) = \frac{\sigma_k(1)}{\sqrt{N}}
        \end{equation}
        with $\sigma_k(1) = \order{1}$, regardless of the absolute size of $S_k$. The latter can best be shown
        for the example of uncertainties of two observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
        We find that the the method of moments yields uncertainties on $\vec{S}$ that are roughly
        \danny{$10\%$} larger than those obtained from maximum-likelihood fits and with the same
        number of events \danny{[cite!]}. However, we wish to note that said fits only produce a limited
        subset of the angular observables, and the statistical error of the fit is expected to increase
        with the number of fit parameters \danny{[cite?]}.
    \item For all toy studies, we produce plots of their skewness and their kurtosis as functions of
        $N$. We find in all cases that the mean skewness and kurtosis converge towards zero.
        \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of the number of simulated events $N$.}
\end{itemize}
All results of our toy studies, as summarised above, show consistently that the joint distribution of the angular
observables converges rapidly towards a multivariate Gaussian distribution. We therefore propose to describe the
resulting distribution as a multivariate Gauss when skewness and kurtosis fall below a given threshold.
The precise target threshold for skewness and kurtosis are subject to the details of the underlying physical analysis, and reside therefore outside the scope of this work.

@@

\section{Conclusion}

We have carried out a combined analytical and numerical study of the method of moments; a method for
the extraction of angular observales from the angular distribution of a generic multi-body process.
We have studied the performance of the method of moments using pseudo data derived from the SM
predctions for one uniangular decay ($B\to K\ell^+\ell^-$) and one triangular decay ($B\to K^*\ell^+\ell^-$).
From this, we find rapid convergence of the joint likelihood of the angular observables towards a multivariate Gaussian.
We draw the conclusion that this method exhibits several benefits in the determination of angular observables when
compared with a maximum-likelihood fit.\\

First, we find no bias in the determinations of the angular observables even for a small number of events.
However, this comes at the expense of a slight increase of the order of $10\%$\todo{check!} for the statistical uncertainty
on the mean.
Second, in the absence of detector effects, a determination using the method of moments is correct and unbiased
even when not all aspects of the physical distribution are modelled. This is explicitly shown for the case of
higher partial waves in multibody final states.
Third, we develop a systematic method for the dertermination of detector effects that lead to
dilution and mixing of the angular observables. We present an algorithm to calculate the necessary
unfolding matrix, which is only computationally feasible when using the method of moments.
Fourth, the fact that the joint distribution of the angular observables that results from the method of moments
is a multivariate Gaussian distribution allows for an easy and precise transfer of correlation
information to subsequent theoretical analyses.
Last but not least, the resulting distribution arises without the need for additional constraints necessary
in order to achieve convergence, such as in some maximum-likelihood fits. Therefore, the results can be
directly combined with similarly-obtained results from different experiments, without the need to remove
the constraints a posteriori.\\

@@

The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).
We would like to thank Ulrik Egede and Konstantinos Petridis for insightful discussions.
\danny{Potential people to read the manuscript: Jochen Dingfelder (Belle-II, Bonn), Ulrik Egede (LHCb, Imperial London), Christoph Langenbruch (LHCb, Warwick), Michel De Cian (LHCb, Heidelberg), Adrian Bevan (ATLAS, Queen Mary)  }
D.v.D would like to thank Gudrun Hiller and Martin Jung for discussing the results of \cite{Das:2014sra} prior to publication.

@@

\section{On the Partial-Wave Expansions of Angular Observables}
\label{app:partial-waves}

@@

From the positivity of amplitudes and unitarity, one immediately finds
\begin{equation}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < \infty\,,
\end{equation}
where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
Using the large-$l$ behaviour of $n_{l,m}$ one finds that with
increasing $l_i$ the partial-wave amplitudes must asymptotically fall
off as
\begin{equation}
    |A_i^{(l_i)}| \Big|_{l_i \gg m_i} \sim \frac{1}{\sqrt{l_i}}\,.\todo{wrong!}
\end{equation}
This motivates to cut off the partial-wave expansion at some arbitrary angular momentum $L$.\\

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of individual contributions, with or without the cutoff $L$.
We close by remarking that for any angular-momentum cutoff $l < L < \infty$, one can find an upper limit for the squared sum of all remaining partial waves through a comparison with the inclusive decay
width.

>>>>

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare decay FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, note that the method we
describe in the following is general, and it applies to arbitrary decay
or scattering processes. We find a previous work \cite{Dighe:1998vk}
that advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays, but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this previous
work by studying the uncertainties that are introduced by mismodelling
of the angular distribution, and by working out a recipe to determine
and unfold detector effects. Moreover, we show
that the method of moments has two major advantages over the usual
approach based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF, see e.g. \cite{Egede:1142152}.
        The method of moments has no such convergence problems.
    \item The results of likelihood fits suffer from systematic uncertainties
        and systematic bias (see e.g. \cite{Egede:1142152}), especially in but not limited
        to cases in which the PDFs do not faithfully model the underlying physics. We will
        show that the method of moments does not suffer a very particular type of
        mismodelling that arises from introducing a cutoff in a partial wave expansion.
    \item Using the method of moments, the joint probability distribution of the
        angular observables converges towards a multivariate Gaussian distribution.
        This allows an easy transfer of correlation information from the experiments to
        interested theorists.
\end{enumerate}

@@

Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the distribution of the $\mathrm{pull}$
        of any observables $S_i$,
        \begin{equation}
            \mathrm{pull_i} \equiv \frac{\est{S}_i - S_i}{\est{\sigma}_{i}}\,.
        \end{equation}
        Here $S_i$ refers to the true (input) value for the angular observables, $\est{S}_i$ refers to the
        mean of the pseudomeasurement via the MC estimate, and $\est{\sigma}_{i} \equiv \sqrt{\est{\Sigma_{ii}}}$
        refers to an estimator of the standard deviation of the pseudomeasurement. All pull distributions
        obtained in our studies can be successfully
        fitted to a gaussian distribution. Due to the large number of studied distributions, we have
        to restrict ourselved to a subset. In our opinion the pull distributions for the observables
        $S_5 \simeq 27\%$ and $S_7 \simeq 2\%$, which are obtained from SM-like $B\to K^*\ell^+\ell^-$
        decays, are good examples. We show their respective pull distributions for $2\cdot 10^5$ toy studies
        of $200$ events per study in \reffig{pulls}.
    \item We study the MC estimate for the absolute uncertainty $\est{\sigma_i(N)}$ with respect to the angular observable $S_i$
        as a function of the number of simulated events $N$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \label{eq:unc-on-mean}
            \est{\sigma_i(N)} = \frac{\sigma_i}{\sqrt{N}}
        \end{equation}
        with $\sigma_i(1) = \order{1}$, regardless of the absolute size of $S_k$. The latter can best be shown
        for the example of uncertainties of two observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
        We find that the the method of moments yields uncertainties on $\vec{S}$ that are roughly
        $10\%$ larger than those obtained from maximum-likelihood fits and for the same
        number of events \danny{[cite!]}. However, we wish to note that said fits only produce a limited
        subset of the angular observables, and the statistical error of the fit is expected to increase
        with the number of fit parameters \danny{[cite?]}. \danny{The method of moments, on the other hand,
        does not suffer from this problem.}
%    \item For all toy studies, we produce plots of their skewness and their kurtosis as functions of
%        $N$. We find in all cases that the mean skewness and kurtosis converge towards zero.
%        \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of the number of simulated events $N$.}
\end{itemize}
All results of our toy studies, as summarised above, show consistently that the joint distribution of the angular
observables converges rapidly towards a multivariate Gaussian distribution. We therefore propose to describe the
resulting distribution as a multivariate Gauss when skewness and kurtosis fall below a given threshold.
The precise target threshold for skewness and kurtosis are subject to the details of the underlying physical analysis, and reside therefore outside the scope of this work.

@@

\section{Summary}

We have carried out a combined analytical and numerical study of the method of moments; a method for
the extraction of angular observales from the angular distribution of a general multi-body process.
We have studied the performance of the method of moments using pseudo data derived from the SM
predctions for one uniangular decay ($B\to K\ell^+\ell^-$) and one triangular decay ($B\to K^*\ell^+\ell^-$).
From this, we find rapid convergence of the joint likelihood of the angular observables towards a multivariate Gaussian.
We draw the conclusion that this method exhibits several benefits in the determination of angular observables when
compared with a maximum-likelihood fit.\\

First, we find no bias in the determinations of the angular observables even for a small number of events.
However, this comes at the expense of a slight increase of the order of $10\%$ for the statistical uncertainty
on the mean.
Second, in the absence of detector effects, a determination using the method of moments is correct and unbiased
even when not all aspects of the physical distribution are modelled. This is explicitly shown for the case of
higher partial waves in multibody final states.
Third, we develop a systematic method for the dertermination of detector effects that lead to
dilution and mixing of the angular observables. We present an algorithm to calculate the necessary
unfolding matrix, which is only computationally feasible when using the method of moments.
Fourth, the fact that the joint distribution of the angular observables that results from the method of moments
is a multivariate Gaussian distribution allows for an easy and precise transfer of correlation
information to subsequent theoretical analyses.
Last but not least, the resulting distribution arises without the need for additional constraints
in order to achieve convergence, such as e.g. in some maximum-likelihood fits. Therefore, results obtained
using the method of moments can be directly averaged, without the need to remove
the constraints a posteriori.\\

@@

We would like to thank Ulrik Egede, Gudrun Hiller, Martin Jung and Konstantinos Petridis for insightful discussions.
The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).

@@

\section{On the Partial-Wave Expansion of Angular Observables}
\label{app:partial-waves}

@@

\danny{
From the orthogonality of the Legendre polynomials, one immediately finds
\begin{equation}
    \label{eq:partial-waves:bound}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < S_\text{incl}\,,
\end{equation}
where $S_\text{incl}$ is the corresponding observable in the associated inclusive decays,
and where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
The positivity of the amplitudes in \refeq{partial-waves:bound} implies that we can estimate
the error introduced by cutting off the expansion at some arbitrary angular momentum $L$. One
obtains
\begin{equation}
    |A_i^{(l' > L)}|^2 < S_\text{incl} - \sum_{l=0}^L |A_i^{(l)}|^2\,.
\end{equation}
}

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of the neglected partial waves, e.g. as outlined above.
